---
title: "Workshop Demonstrating extraChIPs"
author: Stevie Pederson^[Black Ochre Data Labs, The Kids Research Institute Australia stevie.pederson@thekids.org.au]
bibliography: ../inst/references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An Example Workshop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# The package extraChIPs

## Overview

### Description

This is a 90 minute workshop written for EuroBioc26 and a Bioconductor Workflow under active development.
The primary aim is to enable analysis of ChIP-Seq data, and has been developed with this specific data type in mind.
However, the principles can plausibly be extended to ATAC-Seq data, with a clear caveat around the counting of alignments, which may require more care for ATAC-Seq.

Two different ChIP targets (ER$\alpha$ and H3K27ac) will be analysed using complementary strategies, then a data integration stage performed, highlighting the true capabilities of `r BiocStyle::Biocpkg("extraChIPs")`.
Both ChIP targets were prepared in the ZR-75-1 breast cancer cell line under control conditions (Estrogen; E2) then supplemented by *dihydrotestosterone* (E2+DHT).
Raw data is available at https://www.ncbi.nlm.nih.gov/Traces/study/?acc=PRJNA509779&o=acc_s%3Aa with these samples pre-processed using the code at https://github.com/smped/PRJNA509779 [@Pederson-fd].

The original publication containing the detailed biological background is at https://doi.org/10.1038/s41591-020-01168-7 [@Hickey2021-mz]

### Pre-requisites

List any workshop prerequisites, for example:

* Basic knowledge of R syntax
* Understanding of piping in `R` 
* Familiarity with `ggplot2` and the wider `tidyverse`
* Familiarity with the GenomicRanges and SummarizedExperiment classes

<!-- List relevant background reading for the workshop, including any -->
<!-- theoretical background you expect students to have. -->

<!-- * List any textbooks, papers, or other reading that students should be -->
<!--   familiar with. Include direct links where possible. -->

### Participation

<!-- Describe how students will be expected to participate in the workshop. -->

### _R_ / _Bioconductor_ packages used

Whilst the primary focus will be on the usage of `extraChIPs`, additional important packages being discussed will include:

- `r BiocStyle::Biocpkg("qsmooth")` [@Hicks2018-uw]
- `r BiocStyle::Biocpkg("quantro")` [@Hicks2015-ee]
- `r BiocStyle::Biocpkg("plyranges")` [@Lee209-gb]
<!-- - `r BiocStyle::Biocpkg("DiffBind")` and `r BiocStyle::Biocpkg("csaw")` -->

```{r}
library(extraChIPsWorkshop)
library(rtracklayer)
library(tidyverse)
```


### Time outline

| Activity                       | Time |
|--------------------------------|------|
| Annotation Preparation         | 15m  |
| Fixed-Width Peaks (ER$\alpha$) | 30m  |
| Sliding Windows                | 30m  |
| Integration of Datasets        | 15m  |

### Workshop goals and objectives

Today we will be analysing two ChIP-Seq datasets.
All pre-processing has been performed, including read QC, deduplication, alignment to the genome and peak calling.
This will have produced `.bam` files containing alignments and `.narrowPeak` files identifying the genomic regions identified as peaks by MACS2 [@Zhang2008-ms].

### Learning goals

* preparation of genomic region annotations for interpretation of ChIP-seq data
* making informed decisions for normalisation
* performing Differential Signal Analysis (DSA)
* integration of two or more ChIP-seq targets

### Learning objectives

<!-- * analyze xyz data to produce... -->
<!-- * create xyz plots -->
<!-- * evaluate xyz data for artifacts -->

## Workshop

1. Annotation Preparation
2. Differential Signal Analysis Using Fixed-Width Peaks
3. Differential Signal Analysis Using Sliding Windows
4. Integration of two or more ChIP targets

# Annotation Preparation

During sample preparation, a reference genome will have been selected and for today's dataset, reads were aligned to the hg19/GRCh37 reference genome.
This will contain not only chromosome sequences, but a set of annotations will have been defined annotating genes, transcripts and exons.
For today, we'll use those defined by [GENCODE version v44, lifted over to hg19](https://www.gencodegenes.org/human/release_44lift37.html), and which roughly corresponds to Ensembl Release 110.

The genome information can usually be obtained from the header of any associated bam files, but as an initial introduction, `extraChIPs` can provide this easily and flexibly.
The key data structure in the Bioconductor ecosystem for the information is a `Seqinfo` object.

```{r sq}
sq <- defineSeqinfo("GRCh37", chr = TRUE)
sq
```

`defineSeqInfo()` is a quick and easy shortcut for defining a `Seqinfo` object, however no scaffolds or patches will be included, only the main chromosomes.
By default the mitochondrial chromosome will also be omitted, although this can easily be added by passing the argument `mito = "M"` to include `chrM` in the output.
The `chr` prefix is added by default and this can be omitted by passing `chr = FALSE`

Next we might like to load our gene annotations.
These have been provided in the directory `inst/extdata` and contain only the gene, transcript and exon-level annotations from the original gtf.
After loading, add the seqinfo object to the returned GenomicRanges object to ensure all of our objects contain the identical genomic co-ordinates.

```{r}
f <- system.file(
  'extdata', 'gencode.v44lift37.reduced.gtf.gz', package = "extraChIPsWorkshop"
)
gtf <- import.gff(f)
seqinfo(gtf) <- sq
gtf
```

Now we have genes annotated, we know where all transcripts begin and end.
We can use `defineRegions()` to define promoters, upstream promoters, gene-bodies and intragenic regions.
this will classify every base in the genome defined by the `Seqinfo` object into one of these features

```{r}
split_gtf <- split(gtf, gtf$type)
regions <- defineRegions(
  genes = split_gtf$gene, 
  transcripts = split_gtf$transcript, 
  exons = split_gtf$exon, 
  intron = FALSE
)
names(regions)
```

We can compare our ChIP peaks to these regions and can also use them to map our peaks to genes.
Every gene and transcript associated with a region has been included in the mcols element for every region.
These won't be included for distal intergenic regions, but will still be included for the proximal intergenic regions

```{r}
regions$proximal_intergenic
```




# References
